\vspace{-12pt}\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-11-06 paper/paper.tex
DCC 2016 submission.

\title
{\large
\textbf{Burrows-Wheeler transform for terabases}
}

@@

The dramatic decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}. The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are many other projects that are similar or greater in scale. A sequencing machine produces a large number of \emph{reads} (short sequences) that cover the sequenced genome many times over. For a 3~Gbp human genome, the total length of the reads is often 100~Gbp or more.

@@

The \emph{Read Server} project at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis. Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads. Initially, the project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project. After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences. These sequences are stored in 16 \BWT-based indexes \cite{Ferragina2005a} taking a total of 561.5~gigabytes.

Preprocessing large datasets can take days or even weeks. It is often not feasible to rebuild everything when new methods of analysis require new functionalities from the indexes. \BWT-based methods were chosen due to their versatility. A \emph{run-length encoded} \BWT{} compresses repetitive sequence collections quite well \cite{Maekinen2010}, while the similarities to the suffix tree and the suffix array make \BWT-based indexes suitable for a wide variety of \emph{pattern matching} and \emph{sequence analysis} tasks \cite{Ohlebusch2013,Maekinen2015}.

Read Server partitions the unique reads based on the last two bases of the read. Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as we cannot compress the similarities between the reads in different indexes. Reducing the number of indexes would improve both memory usage and query performance. However, the lack of \BWT{} construction algorithms that work with terabases of sequence data has made it infeasible so far.

@@

The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}. Given the sequences and their suffix array, we can easily build the \BWT{} of the sequences. These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves. Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
% DCC:
\cite{Kaerkkaeinen2015a}.
% arXiv:
%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.

There are algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first. Some require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}. These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets. \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs. Some algorithms can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming a bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}. GPU-based algorithms can be much faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}. Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, but such algorithms generally use the hardware less efficiently than those running on a single system \cite{Wang2015}.

@@

If suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$. If $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT{} and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order. This forms the essence of \LF\emph{\nobreakdash-mapping}. Let $\mC[c]$ bet the number of suffixes starting with a character smaller than $c$, and let $S.\mrank(i,c)$ be the number of occurrences of $c$ in the prefix $S[1,i]$. Then $\mSA[\mLF(i)]$, where
$$
\mLF(i) = \mLF(i,\mBWT[i])
\,\, \textrm{and} \,\,
\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
$$
points to the suffix $T[\mSA[i]-1,n]$. The more general version of \LF-mapping $\mLF(i,c)$ tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$. This is the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of $cT[\mSA[i],n]$ among the suffixes of text $T$.

The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures. The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found by using \emph{backward searching}. Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern. We find $[sp_{i-1},ep_{i-1}]$ as $[\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])]$. By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT. In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding of the \BWT{}.

@@

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$ for $1 \le j < \abs{T_{i}}$. As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.

@@

Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character \cite{Hon2007}. We find the pointer $\mSA[i]$ to the beginning of text $T$ (where $\mBWT[i] = 0$). Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$. Finally \emph{replace} the endmarker at $\mBWT[i]$ with the inserted character $c$ and \emph{insert} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.

We can use the transformation for \BWT{} construction in several ways. We can do \emph{batch updates} and transform the \BWT{} of text $T$ into that of text $XT$, where $X$ is a string \cite{Hon2007}. We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}. We can also \emph{extend} multiple texts at once by inserting a new character to the beginning of each of them \cite{Bauer2013}. In all cases, we can use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT. Dynamic representations increase the size of the \BWT{} (e.g.~by around 1.5x in \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.

@@

\smallbreak\noindent\textbf{Search.} We search for all texts of collection $\Bcoll$ in $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$. This takes $\Oh(n_{\Bcoll} t_{r})$ time. If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.

@@

If we split a text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections. This takes $\Oh((p+t_{r})n)$ time and uses essentially $\min(n, \frac{n}{p} \log n)$ bits of working space.

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection.
% arXiv:
%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
%
We can reduce the overhead by writing the lexicographic ranks to \emph{disk} in the search phase. If we can sort the ranks on disk, we just need to scan the rank array once during the merge phase. As fast disk space is often scarce, we may want to \emph{interleave} the sorting with the search and merge phases, and to \emph{compress} the ranks before writing them to disk. We now describe the key ideas that make \BWT{} merging fast and space-efficient.

@@

\smallbreak\noindent\textbf{Merge.} The merge phase starts with the rank array stored in multiple files on disk. Each of the files is sorted and compressed. For interleaving the \BWT{}s, we need to merge the files to scan through the rank array. We can also use multiple threads here. One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks. Another thread consumes the stream and uses it to interleave the \BWT{}s. If the disk is fast enough, we may want to use multiple threads for the multiway merge.

@@

We have implemented the improved merging algorithm in a tool for merging the \BWT{}s of large collections of short reads. The tool, \BWTmerge{}, is written in C++, and the source code is available on GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation uses the \emph{SDSL library} \cite{Gog2014b} and the new features in C++11. As a result, it needs a fairly recent C++ compiler to compile. We have successfully built \BWTmerge{} on Linux and OS~X using g++.

The target environment of \BWTmerge{} is a \emph{single node} of a \emph{computer cluster}. The system should have tens of CPU cores and hundreds of gigabytes of memory. There should also be a few hundred gigabytes of reliable (i.e.~local) disk space for temporary files. The number of search threads is equal to the number of CPU cores, while the merge phase uses just one producer thread and one consumer thread. By adjusting the sizes of run buffers and thread buffers and the number of merge buffers, \BWTmerge{} should work reasonably well in different environments.

@@

For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, ensuring that no run continues from one block to the next one. For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$. Each of these seven increasing sequences is encoded using the \emph{sdarray} encoding \cite{Okanohara2007}. To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block. A \select{} query on the same sequence transforms the position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block. Finally, we decompress the block to answer the query. \select{} queries and random accesses to the \BWT{} work in a similar way. There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a relatively short lexicographic range.

We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}. This reduced the space overhead by tens of gigabytes over using \texttt{malloc()} and \texttt{free()}.
% arXiv:
%When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB. Each thread tries to reuse the heap it used for the last allocation. If a heap is full or the thread cannot acquire the mutex, it moves to the next heap. With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle. Most \texttt{free()} calls did not release the memory back to the operating system, while it took a while before any thread could reuse the holes created by the \texttt{free()} calls.

@@

We used a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory. The system was running Ubuntu 12.04 on Linux kernel 3.2.0. We used a development version of \BWTmerge{}  equivalent to v0.3. For the other tools, we used the versions that were available on GitHub in October~2015. All software was compiled with gcc/g++ version 4.9.2. We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files. \BWTmerge{} was allowed to use 32 threads, while the other \BWT{} construction tools were limited to 4 or 5 threads by design. The reported merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers extracted from the human reference genome.

@@

We used two datasets from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}. \CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892). We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459). For each individual, we concatenated the files and corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}). \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned by the last two bases. See Table~\ref{table:datasets} for further details on the datasets.

@@

The results can be seen in Figure~\ref{fig:benchmark} (left). The average speed for inserting 1.06~Tbp into file AA ranged from 8.27 to 9.40~Mbp/s, depending on the parameter values. Memory overhead was 21.1 to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
% arXiv:
%The temporary files used 287 to 306~gigabytes of disk space. Thread buffer size and the number of merge buffers were the most important parameters. The larger the individual files are, the more memory the search phase uses and the faster the merge phase is. Increasing run buffer size to 256~MB made the search phase faster with 512~MB thread buffers and slower with 256~MB thread buffers.
%
For the further experiments, we chose 128~MB run buffers, 256~MB thread buffers, and 6 merge buffers (overhead 30.8~GB).

\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared \BWTmerge{} to the fastest \BWT{} construction tools on general hardware \cite{Li2014a}. We built the \BWT{} of the \CEU{} dataset using \ropebwt{} \cite{Li2011-2013} with parameters \texttt{-btORf -abcr} and \ropebwtii{} \cite{Li2014a} with parameters \texttt{-bRm10g}. We also built individual \BWT{}s using \ropebwt{} and merged them using \BWTmerge. All tools wrote the \BWT{} files in their preferred formats.

The results can be seen in Figure~\ref{fig:benchmark} (right). When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than building the final \BWT{} directly with \ropebwt. \ropebwt2{} ran out of memory just before finishing. \BWTmerge{} was about 1.2x slower and 1.5x more space-efficient than it. The running time of \BWTmerge{} was split almost evenly between building the individual \BWT{}s and merging them.
% arXiv:
%
When \ropebwt{} and \ropebwt2{} sort the reads in \emph{reverse lexicographic order} (RLO, parameter \texttt{-s}) to improve compression, all tools improve their performance. \BWTmerge{} is now 1.70x slower and 2.12x more space-efficient than \ropebwt{} and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2.
% arXiv:
%Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them. Note that \BWTmerge{} builds a \BWT{} for the concatenation of three input files that are in RLO, while the \BWT{}s produced by the other tools are completely in RLO. Maintaining RLO during merging would reduce the size of the final \BWT{} from 58.9~GB to 54.4~GB.

\smallbreak\noindent\textbf{Read Server.} In the last experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC into the first file; TT, GT, CT, AT, TG, GG, CG, AG into the second one). Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~GB and 219~GB of memory, and used 297~GB and 300~GB of disk space, respectively. This reduced the size of the FM-indexes from around 560~GB to 480~GB. By converting the \BWT{}s to the native format of \BWTmerge{}, we further reduced the size of the indexes to 360~GB. This makes it possible to host the indexes on two servers instead of the original three.

@@

\BWTmerge{} can be used as a building block of a \BWT{} construction algorithm. We split the read collection into multiple subcollections, build the \BWT{}s of the subcollections with other tools, and merge the results. The resulting algorithm is typically slower but more space-efficient than the existing algorithms.

>>>>

\title
{\vspace{-12pt}\large
\textbf{Burrows-Wheeler transform for terabases}
}

@@

The decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}. The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are other projects that are similar or greater in scale. A sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over. For a 3~Gbp human genome, the total length of the reads is often 100~Gbp or more.

@@

Preprocessing large datasets can take weeks. It is often not feasible to rebuild everything when new methods of analysis require new functionalities. Structures based on the \emph{Burrows-Wheeler transform} (\BWT) are often useful due to their versatility. A \emph{run-length encoded} \BWT{} compresses repetitive sequence collections quite well \cite{Maekinen2010}, while the similarities to the suffix tree and the suffix array make \BWT-based indexes suitable for many \emph{pattern matching} and \emph{sequence analysis} tasks \cite{Ohlebusch2013,Maekinen2015}.

The \emph{Read Server} project at the Sanger Institute develops tools for large-scale \emph{reference-free} genome analysis, avoiding reference bias. Unique reads are compressed and indexed using the \BWT, while metadata databases contain information on the original reads. Initially, the project works with
% arXiv:
%the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
% DCC:
data from the 1000 Genomes Project.
After error correction and trimming the reads to either 73~bp or 100~bp, the 922~billion original reads (86~Tbp) are reduced to 53.0~billion unique sequences (4.88~Tbp). These sequences are stored in 16 \BWT-based indexes \cite{Ferragina2005a} taking a total of 561.5~gigabytes.

The unique reads are partitioned between the \BWT{}s by the last two bases. Every query must be repeated in all 16 indexes. The \BWT{}s also require more space, as we cannot compress the similarities between the reads in different indexes. Reducing the number of indexes would improve both memory usage and query performance. This requires \BWT{} construction algorithms that can handle terabases of data.

@@

The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and to derive the \BWT{} from the suffix array. These algorithms cannot be used with large datasets, as they require much more memory than the sequences themselves. Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
% DCC:
\cite{Kaerkkaeinen2015a}.
% arXiv:
%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.

There are many \emph{direct} \BWT{} construction algorithms that do not need the suffix array. Some require a limited amount of working space on top of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, while others use the disk as additional working space \cite{Ferragina2012,Beller2013}. These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s. \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs. Some can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming the bottleneck after about 1~Tbp \cite{Bauer2013,Li2014a}. GPU-based algorithms are even faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}. Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, at the price of using more resources for the construction \cite{Wang2015}.

@@

If $X \le Y$ in lexicographic order, we also have $cX \le cY$ for any character $c$. If $\mBWT[i] = c$ is the $j$\nobreakdash-th occurrence of $c$ in the \BWT{} and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.

Let $\mC[c]$ bet the number of suffixes starting with a character smaller than $c$, and let $S.\mrank(i,c)$ be the number of occurrences of $c$ in the prefix $S[1,i]$. We define \LF-\emph{mapping} as $\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c)$ and $\mLF(i) = \mLF(i,\mBWT[i])$. The general form $\mLF(i,c)$ is the number of suffixes $X$ of text $T$ with $X \le cT[\mSA[i],n]$. This is known as the  \emph{lexicographic rank} $\mrank(cT[\mSA[i],n],T)$ of text $cT[\mSA[i],n]$ among the suffixes of text $T$. The specific form $\mLF(i)$ gives the lexicographic rank of the previous suffix ($\mSA[\mLF(i)] = \mSA[i]-1$, if $\mSA[i] > 1$, and $\mSA[\mLF(i)] = n$ otherwise).

The \emph{FM-index} (\FMI) \cite{Ferragina2005a} is a full-text index based on the \BWT{}. We use \emph{backward searching} in the FM-index to find the lexicographic range $[sp,ep]$ matching pattern $P$. Let $[sp_{i},ep_{i}]$ be the range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern. We find $[sp_{i-1},ep_{i-1}]$ as $[\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])]$. By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT. In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding of the \BWT{}.

@@

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$, for $1 \le j < \abs{T_{i}}$. As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.

@@

Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character \cite{Hon2007}. We find the pointer $\mSA[i]$ to the beginning of text $T$ (where $\mBWT[i] = 0$). Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$. Finally we \emph{replace} the endmarker at $\mBWT[i]$ with the inserted character $c$ and \emph{insert} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.

We can use the transformation for \BWT{} construction in several ways. We can use \emph{batch updates} and transform the \BWT{} of text $T$ into that of text $XT$, where $X$ is a string \cite{Hon2007}. We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}. We can also \emph{extend} multiple texts at once by inserting a new character to the beginning of each of them \cite{Bauer2013}. In all cases, we can use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT. Dynamic representations increase the size of the \BWT{} (e.g.~by around 1.5x in \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.

@@

\smallbreak\noindent\textbf{Search.} We search for all texts of collection $\Bcoll$ in $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of $\Bcoll$. This takes $\Oh(n_{\Bcoll} t_{r})$ time. We either need the collection in plain form, or extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.

@@

If we split a text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections. This takes $\Oh((p+t_{r})n)$ time and uses essentially $\min(\frac{n}{p} \log n, n)$ bits of working space.

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{}.
% arXiv:
%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
%
We can reduce the overhead by writing the lexicographic ranks to \emph{disk}. If we sort the ranks on disk, we just need to scan the rank array once during the merge phase. We can also \emph{compress} the ranks before writing them to disk and \emph{interleave} the sorting with the search and merge phases. We now describe the key ideas for fast and space-efficient \BWT{}.

@@

\smallbreak\noindent\textbf{Merge.} The ranks are stored in sorted order in multiple files on disk. For interleaving the \BWT{}s, we need to merge the files and to scan through the rank array. We can also use multiple threads here. One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks. Another thread consumes the stream and uses it to interleave the \BWT{}s. If the disk is fast enough, we may want to use multiple threads for the multiway merge.

@@

We have implemented the improved merging algorithm as a tool for merging the \BWT{}s of large read collections. The tool, \BWTmerge{}, is written in C++, and the source code is available on GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation uses the \emph{SDSL library} \cite{Gog2014b} and the new features in C++11. As a result, it needs a fairly recent C++ compiler to compile. We have successfully built \BWTmerge{} on Linux and OS~X using g++.

The target environment of \BWTmerge{} is a \emph{single node} of a \emph{computer cluster}. The system should have tens of CPU cores, hundreds of gigabytes of memory, and hundreds of gigabytes of local disk space for temporary files. The number of search threads is equal to the number of CPU cores, while the merge phase uses just one producer thread and one consumer thread. \BWTmerge{} can be adapted to many other environments by adjusting the number and the size of the buffers.

@@

For \rank/\select{} support, we divide the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, ensuring that the runs do not cross block boundaries. For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$. These increasing sequences are stored using the \emph{sdarray} encoding \cite{Okanohara2007}. To compute $\mBWT.\mrank(j,c)$, we start with a \rank{} query on the $n_{i}$ sequence to find the block. A \select{} query on the same sequence transforms $j$ into a block offset, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block. We then decompress the block to answer the query. \select{} queries and accessing the \BWT{} work in a similar way. There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a short lexicographic range.

We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}. This reduces the space overhead by tens of gigabytes over using \texttt{malloc()} and \texttt{free()}.
% arXiv:
%When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB. Each thread tries to reuse the heap it used for the last allocation. If a heap is full or the thread cannot acquire the mutex, it moves to the next heap. With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle. Most \texttt{free()} calls did not release the memory back to the operating system, while it took a while before any thread could reuse the holes created by the \texttt{free()} calls.

@@

We used a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory. The system was running Ubuntu 12.04 on Linux kernel 3.2.0. We used a development version of \BWTmerge{}  equivalent to v0.3, and the versions of the other tools that were available on GitHub in October 2015. All software was compiled with gcc/g++ version 4.9.2. We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB disk for temporary files. \BWTmerge{} used 32 threads, while the other \BWT{} construction tools were limited by design to 4 or 5 threads.
% arXiv:
%The merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers.

@@

Our datasets come from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}. \CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892). We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459). For each individual, we concatenated the files and corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}). \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3. There are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built by using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned by the last two bases. See Table~\ref{table:datasets} for further details on the datasets.

@@

The results can be seen in Figure~\ref{fig:benchmark} (left). The average speed for inserting 1.06~Tbp into file AA ranged from 8.27~Mbp/s to 9.40~Mbp/s, depending on the parameter values. Memory overhead was 21.1~GB to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
% arXiv:
%The temporary files used 287 to 306~gigabytes of disk space. Thread buffer size and the number of merge buffers were the most important parameters. The larger the individual files are, the more memory the search phase uses and the faster the merge phase is. Increasing run buffer size to 256~MB made the search phase faster with 512~MB thread buffers and slower with 256~MB thread buffers.
%
For the further experiments, we chose 128~MB run buffers, 256~MB thread buffers, and 6 merge buffers (overhead 30.8~GB).

\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared \BWTmerge{} to the fastest \BWT{} construction tools on general hardware \cite{Li2014a}. We built the \BWT{} of the \CEU{} dataset using \ropebwt{} \cite{Li2011-2013} with parameters \texttt{-btORf -abcr} and \ropebwtii{} \cite{Li2014a} with parameters \texttt{-bRm10g}. We also built individual \BWT{}s using \ropebwt{} and merged them with \BWTmerge. All tools were set to write the \BWT{}s in their preferred formats.

The results are in Figure~\ref{fig:benchmark} (right). When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than \ropebwt. \ropebwt2{} ran out of memory just before finishing. It would have been about 1.2x faster and 1.5x less space-efficient than \BWTmerge. The running time of \BWTmerge{} was split evenly between \BWT{} construction and merging.
% arXiv:
%
When \ropebwt{} and \ropebwt2{} sort the reads in \emph{reverse lexicographic order} (RLO) to improve compression, all tools improve their performance. \BWTmerge{} becomes 1.70x slower and 2.12x more space-efficient than \ropebwt{}, and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2.
% arXiv:
%Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them. Note that \BWTmerge{} builds a \BWT{} for the concatenation of three input files that are in RLO, while the \BWT{}s produced by the other tools are completely in RLO. Maintaining RLO during merging would reduce the size of the final \BWT{} from 58.9~GB to 54.4~GB.

\smallbreak\noindent\textbf{Read Server.} In the last experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC into the first file; TT, GT, CT, AT, TG, GG, CG, AG into the second one). Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~GB and 219~GB of memory, and used 297~GB and 300~GB of disk space, respectively. This reduced the size of the FM-indexes from around 560~GB to 480~GB. By converting the \BWT{}s to the native format of \BWTmerge{}, we further reduced the size of the indexes to 360~GB.
% arXiv:
%This makes it possible to host the indexes on two servers instead of the original three.

@@

\BWTmerge{} can be used as a part of a \BWT{} construction algorithm. We split the read collection into subcollections, build the \BWT{}s of the subcollections, and merge the results. The resulting algorithm is typically slower but more space-efficient than the existing algorithms.
